### 2023-10-04 doc/_bayes_freq.qmd
small fixes
several changes, including Griffin's suggestions

## Defining the Bayesian model

>>>>

## Defining the Bayesian model {#sec-bayes_def}

In the main text, we specify the inference problem we must solve as being of the
form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-full_inference_SI}
Here, we briefly define the missing nuisance parameters. Let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires cycle numbers $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
frequency values. As with @eq-R-mat_split in the main text, we can split
$\underline{\underline{F}}$ into two matrices of the form
$$
\underline{\underline{F}} = \left[ 
\underline{\underline{F}}^N \; \underline{\underline{F}}^M
\right],
$${#eq-F-mat_split}
to separate the corresponding neutral and non-neutral barcode frequencies.

### 2023-10-04 doc/_bayesian_inference.qmd
several changes, including Griffin's suggestions

Our objective is to compute the joint probability distribution for all relative
fitness values given our data. We can express this joint posterior distribution
using Bayes theorem as
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines a probability density function, unless
otherwise stated. When defining our statistical model, we need not to focus on
the denominator on the right-hand side of @eq-bayes_obj. Thus, we can write
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M).
$${#eq-bayes_obj_propto}
However, when implementing the model computationally, the normalization constant
on the right-hand side of @eq-bayes_obj must be computed. This can be done from
the definition of the model via an integral of the form
$$
\pi(\underline{\underline{R}}) = \int d^M \underline{s}^M
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M),
$${#eq-bayes_evidence}
also known as a marginalization integral. Hereafter, differentials of the form
$d^n$ imply a $n$-dimensional integral.

Although @eq-bayes_obj and @eq-bayes_obj_propto seem simple enough, recall that
@eq-fitness relates barcode frequency values and the population mean fitness to
the mutant relative fitness. Therefore, we must include these nuisance
parameters as part of our inference problem. To include these nuisance
parameters, let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires cycle numbers $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
frequency values. As with @eq-R-mat_split, we can split
$\underline{\underline{F}}$ into two matrices of the form
$$
\underline{\underline{F}} = \left[ 
\underline{\underline{F}}^N \; \underline{\underline{F}}^M
\right],
$${#eq-F-mat_split}
to separate the corresponding neutral and non-neutral barcode frequencies. With
these nuisance variables in hand, the full inference problem we must solve takes
the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
To recover the marginal distribution over the non-neutral barcodes relative
fitness values, we can numerically integrate out all nuisance parameters, i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}

@@

To overcome this computational limitation, in this work, we rely on the recent
development of the automatic differentiation variational inference algorithm
(ADVI) [@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =
(\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
replaced by an approximate posterior distribution $q_\phi(\theta)$, where $\phi$
fully parametrizes the approximate distribution. As further explained in
@sec-vi_primer, the numerical integration problem is replaced by an optimization
problem of the form
$$
q^*_\phi(\theta) = \min _\phi
D_{KL}(
    q_\phi(\theta) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
$${#eq-vi_objective}
where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the
complicated high-dimensional numerical integration problem is transformed into a
much simpler problem of finding the value of the parameters $\phi$ such that
@eq-vi_objective is satisfied as best as possible within some finite computation
time. Although to compute @eq-vi_objective, we require the posterior
distribution we are trying to approximate $\pi(\theta \mid
\underline{\underline{R}})$, it can be shown that maximizing the so-called
evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the
variational free energy [@buckley2017]---is mathematically equivalent to
performing the optimization prescribed by @eq-vi_objective. We direct the reader
to @sec-vi_primer for a short primer on variational inference.

>>>>

Our objective is to compute the joint probability distribution for all relative
fitness values given our data. We can express this joint posterior distribution
using Bayes theorem as
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines a probability density function. When
defining our statistical model, we need not to focus on the denominator on the
right-hand side of @eq-bayes_obj. Thus, we can write
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M).
$${#eq-bayes_obj_propto}
However, when implementing the model computationally, the normalization constant
on the right-hand side of @eq-bayes_obj must be computed. This can be done from
the definition of the model via an integral of the form
$$
\pi(\underline{\underline{R}}) = \int d^M \underline{s}^M
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M),
$${#eq-bayes_evidence}
also known as a marginalization integral. Hereafter, differentials of the form
$d^n$ imply a $n$-dimensional integral.

Although @eq-bayes_obj and @eq-bayes_obj_propto seem simple enough, recall that
@eq-fitness relates barcode frequency values and the population mean fitness to
the mutant relative fitness. Therefore, we must include these nuisance
parameters as part of our inference problem. We direct the reader to the
supplementary materials for the exact definitions of these parameters. Here, it
suffices to say that the inference problem must include the vector
$\underline{\bar{s}}_T$ of all population mean fitness values and the matrix
$\underline{\underline{F}}$ of all barcode frequencies within the sequencing
data. With these nuisance variables in hand, the full inference problem we must
solve takes the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
To recover the marginal distribution over the non-neutral barcodes relative
fitness values, we can numerically integrate out all nuisance parameters, i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}

@@

To overcome this computational limitation, we rely on the recent development of
the automatic differentiation variational inference algorithm (ADVI)
[@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =
(\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
replaced by an approximate posterior distribution $q_\phi(\theta)$, where $\phi$
fully parametrizes the approximate distribution. As further explained in
@sec-vi_primer, the numerical integration problem is replaced by an optimization
problem of the form
$$
q^*_\phi(\theta) = \min _\phi
D_{KL}(
    q_\phi(\theta) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
$${#eq-vi_objective}
where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the
complicated high-dimensional numerical integration problem is transformed into a
much simpler problem of finding the value of the parameters $\phi$ such that
@eq-vi_objective is satisfied as best as possible within some finite computation
time. Although to compute @eq-vi_objective, we require the posterior
distribution we are trying to approximate $\pi(\theta \mid
\underline{\underline{R}})$, it can be shown that maximizing the so-called
evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the
variational free energy [@buckley2017]---is mathematically equivalent to
performing the optimization prescribed by @eq-vi_objective. We direct the reader
to @sec-vi_primer for a short primer on variational inference.

### 2023-10-04 doc/_experiment.qmd
several changes, including Griffin's suggestions

To determine the relative fitness of the ensemble of genotypes, a series of
growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time; then, an
aliquot is inoculated into fresh media for the next growth cycle. This process
is repeated for roughly 4-7 cycles, depending on the initial abundances of the
mutants and their relative growth rates. The DNA barcodes are sequenced at the
end of each growth cycle to quantify the relative abundance of each of the
barcodes. We point the reader to @kinsler2020 for specific details on these
assays for *S. cerevisiae* and to @ascensao2023 for equivalent assays for *E.
coli*. @fig-01(B) presents a typical barcode trajectory where the darker barcode
trajectories represent the so-called *neutral lineages*, genetically equivalent
to the untagged ancestor strain that initially dominates the culture. These
spiked-in neutral lineages simplify the inference problem since the fitness
metric of all relevant barcodes is quantified with respect to these
barcodes---thus referred to as *relative fitness*.

>>>>

To determine the relative fitness of the ensemble of genotypes, a series of
growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time; then, an
aliquot is inoculated into fresh media for the next growth cycle. This process
is repeated for roughly 4-7 cycles, depending on the initial abundances of the
mutants and their relative growth rates. The DNA barcodes are sequenced at the
end of each growth cycle to quantify the relative abundance of each of the
barcodes. We point the reader to @kinsler2020 for specific details on these
assays for *S. cerevisiae* and to @ascensao2023 for equivalent assays for *E.
coli*. @fig-01(B) presents a typical barcode trajectory where the black
trajectories represent the so-called *neutral lineages*, genetically equivalent
to the untagged ancestor strain that initially dominates the culture. These
spiked-in neutral lineages simplify the inference problem since the fitness
metric of all relevant barcodes is quantified with respect to these
barcodes---thus referred to as *relative fitness*.

### 2023-10-04 doc/_hierarchical_replicates_results.qmd
several changes, including Griffin's suggestions

Our inference pipeline can be extended to account for multiple experimental
replicates via Bayesian hierarchical models [@betancourt2013]. Briefly, when
accounting for multiple repeated measurements of the same phenomena, there are
two extreme cases one can use to perform the data analysis: On the one hand, we
can treat each measurement as entirely independent, losing the power to utilize
multiple measurements when trying to learn a single parameter. This can hurt the
inference since, in principle, the value of our parameter of interest should not
depend on the particular experimental replicate in question. However, this
approach does not allow us to properly "combine" the uncertainties in both
experiments when performing the inference. On the other hand, we can pool all
data together and treat our different experiments as a single measurement with
higher coverage. This loses the subtle differences from biotic and abiotic batch
effects, effectively halving the data that goes into our inference problem.

>>>>

Our inference pipeline can be extended to account for multiple experimental
replicates via Bayesian hierarchical models [@betancourt2013]. Briefly, when
accounting for multiple repeated measurements of the same phenomena, there are
two extreme cases one can use to perform the data analysis: On the one hand, we
can treat each measurement as entirely independent, losing the power to utilize
multiple measurements when trying to learn a single parameter. This can
negatively impact the inference since, in principle, the value of our parameter
of interest should not depend on the particular experimental replicate in
question. However, this approach does not allow us to properly "combine" the
uncertainties in both experiments when performing the inference. On the other
hand, we can pool all data together and treat our different experiments as a
single measurement with higher coverage. This loses the subtle differences due
to biotic and abiotic batch effects, effectively halving the data that goes into
our inference problem.

### 2023-10-04 doc/_single_results.qmd
several changes, including Griffin's suggestions

To visualize the performance of our inference pipeline in fitting our fitness
model to the observed data, we compute the so-called posterior predictive checks
(PPC). In short, the PPC consists of repeatedly generating synthetic datasets in
agreement with the results from the inference results. In other words, we use
the resulting parameter values from the ADVI inference to generate possible
datasets in agreement with the inferred values (See @sec-ppc for further details
on these computations). @fig-02(C) shows these results for all neutral lineages
(upper left corner plot) and a few representative non-neutral barcodes. The
different color shades represent the 95\%, 68\%, and 5\% credible regions, i.e.,
the regions where we expect to find the data with the corresponding
probability---or in terms of our parameter, the $X\%$ credible region is the
interval where we expect the true parameter to lie with $X\%$ probability.

The main advantage of our method is this natural interpretability of these
credible regions where an $X\%$ credible region indeed captures the region of
parameter space where we expect with $X\%$ probability the actual value of the
parameter lies given our statistical model, our prior information, and the
observed experimental data. A common mistake in the literature is interpreting
frequentist confidence intervals as Bayesian credible regions when they are not
equivalent [@morey2016]. Frequentist confidence intervals and Bayesian credible
regions are based on fundamentally different philosophical approaches to
statistics. Frequentist confidence intervals represent the range of values that
would contain the true population parameter with a certain probability if the
experiment was repeated many times. The confidence level does not represent the
probability that the interval contains the true value. According to a specific
model and prior information, Bayesian credible regions represent the range of
values that contain the parameter with a certain posterior probability. The
credible region directly represents the probability that the region contains the
true value. So, frequentist confidence intervals cannot be interpreted as
Bayesian credible regions because they have fundamentally different meanings.
Treating an $X\%$ confidence interval like an $X\%$  credible region is
fallacious since confidence levels do not represent probabilistic coverage of
the true value like credible regions. The intervals are generated through
entirely different procedures.

>>>>

To visualize the performance of our inference pipeline in fitting our fitness
model to the observed data, we compute the so-called posterior predictive checks
(PPC). In short, the PPC consists of repeatedly generating synthetic datasets in
agreement with the results from the inference results. In other words, we use
the resulting parameter values from the ADVI inference to generate possible
datasets in agreement with the inferred values (See @sec-ppc for further details
on these computations). @fig-02(C) shows these results for all neutral lineages
(upper left corner plot) and a few representative non-neutral barcodes. The
different color shades represent the 95\%, 68\%, and 5\% credible regions, i.e.,
the regions where we expect to find the data with the corresponding
probability---or in terms of our parameter, the $X\%$ credible region is the
interval where we expect the true parameter value to lie with $X\%$ probability.

The main advantage of our method is this natural interpretability of these
credible regions where an $X\%$ credible region indeed captures the region of
parameter space where we expect with $X\%$ probability the actual value of the
parameter lies given our statistical model, our prior information, and the
observed experimental data. A common mistake in the literature is interpreting
frequentist confidence intervals as Bayesian credible regions when they are not
equivalent [@morey2016]. Frequentist confidence intervals and Bayesian credible
regions are based on fundamentally different philosophical approaches to
statistics. Frequentist confidence intervals represent the range of values that
would contain the true population parameter with a certain probability if the
experiment was repeated many times. The confidence interval does not represent
the probability that the interval contains the true value. According to a
specific model and prior information, Bayesian credible regions represent the
range of values that contain the parameter with a certain posterior probability.
The credible region directly represents the probability that the region contains
the true value. So, frequentist confidence intervals cannot be interpreted as
Bayesian credible regions because they have fundamentally different meanings.
Treating an $X\%$ confidence interval like an $X\%$  credible region is
fallacious since confidence intervals do not represent probabilistic coverage of
the true value like credible regions. The intervals are generated through
entirely different procedures.
